---
title: "PSTAT 131 Final Project"
author: "Maxwell Landstrom, Shivum Ganesh"
format: pdf
editor: visual
---

[Copied from instructions:]{.underline} Predicting voter behavior is complicated for many reasons despite the tremendous effort in collecting, cleaning, analyzing, and understanding many available datasets. For this final project, we will visualize and analyze the 2024 U.S. presidential election. In addition, we will combine the Untied States county-level census data with the election data. Our target would then be building and selecting classification models (among many predictive models that we’ve covered in this quarter) to predict the election winner.

## Data

```{r}
library(tidyverse)
## read election data 
election_raw <- read_csv("election_raw.csv", col_names = TRUE) 
## read census data
census <- read_csv("census.csv") 
```

1.  election_raw

```{r}
print("dimension of election_raw:")
dim(election_raw)
print("missing value count:")
sum(is.na(election_raw))
print("distinct 'state' values:")
n_distinct(election_raw$State)
```

2.  census

```{r}
print("dimension of census dataframe:")
dim(census)
print("missing value count:")
sum(is.na(census))
print("distinct 'county' values:")
n_distinct(census$County)
```

We computed 1960 distinct county values in the census dataframe. However, election_raw reports 3160 counties in election_raw. This is because many states share county

## Data Wrangling

3.  Aggregated dataset

```{r}
library(dplyr)
election_state <- election_raw%>%group_by(State)%>%summarise(total_gop=sum(votes_gop), total_dem=sum(votes_dem), .groups = 'drop')
```

4.  State and County Winners

```{r}
county_winner <- election_raw%>%
  group_by(State,County)%>%
  summarise(
    total_gop=sum(votes_gop),
    total_dem=sum(votes_dem),
    winner=ifelse(total_gop>total_dem, "gop","dem"),
    FIPS = FIPS,
    .groups = 'drop'
  )
state_winner <- election_raw%>%
  group_by(State)%>%
  summarise(
    total_gop=sum(votes_gop),
    total_dem=sum(votes_dem),
    winner=ifelse(total_gop>total_dem, "gop","dem"),
    .groups = 'drop'
  )

```

## Visualization

[Copied from instructions:]{.underline} Visualization is crucial for gaining insight and intuition during data mining. We will map our data onto maps. In this part, we will work towards building state/county-level red/blue map plots that are commonly shown on [media coverage](#0).

5.  county map

```{r}
library(usmap)
states <- usmap::us_map(regions="counties")

ggplot(states) + 
  geom_sf(aes(fill=county), color = "white") + 
  coord_sf() +
  guides(fill = "none")
```

6.  state map

```{r}
# state-level plotting data
states <- states%>%
  filter(abbr != "PR")%>%
  group_by(full)%>%
  summarise(
    .groups = 'drop'
  )

states <- left_join(states, state_winner, by=c("full"="State"))

```

```{r}
ggplot(states) + 
  geom_sf(aes(fill=winner), color = "white") + 
  coord_sf() +
  guides(fill = "none")
```

7.  California

```{r}
ca_counties <- usmap::us_map(regions="counties", include = "CA")
ca_winner <- county_winner%>%
  filter(State == "California")

ca_counties <- left_join(ca_counties, ca_winner, by=c("county"="County"))

ggplot(ca_counties) + 
  geom_sf(aes(fill=winner), color = "white") + 
  coord_sf() +
  guides(fill = "none")
```

8.  Open-ended

Vote share gradient for Democratic vs Republican.

```{r}
ca_county_data <- ca_winner%>%
  mutate(total_votes = total_gop+total_dem,
         vote_share = total_dem/total_votes)
```

```{r}
ca_counties <- left_join(ca_counties, ca_county_data, by = c("county"="County"))
ggplot(ca_counties) + 
  geom_sf(aes(fill=vote_share), color = "white") + 
  coord_sf() +
  scale_fill_gradient2(
    low = "red",
    mid = 'grey',
    high = 'blue',
    midpoint = 0.5,
    name = "Vote Share",
    labels = scales::percent
  ) +
  guides(fill = "none")
```

9.  Clean county-level census data

```{r}
census_clean <- census%>%
  na.omit() %>%
  mutate(
    pct_male = `Male Population`/`Total Population`,
    pct_female = `Female Population`/`Total Population`,
    pct_whitealone = `White Alone`/`Total Population`,
    pct_black = `Black or African American Alone`/`Total Population`,
    pct_hispanic = `Hispanic or Latino`/`Total Population`,
    Urban = factor(ifelse(`Urban-Rural`=="Urban", 1,0))
  ) %>%
  select(
    County,
    State,
    Urban,
    `Median Household Income`,
    pct_male,
    pct_female,
    pct_whitealone,
    pct_black,
    pct_hispanic
  )
head(census_clean)
```

10. Run PCA for the cleaned county level census data

```{{r}}
pca_data <- census_clean %>%
  select(-County, -State)

pca_data$Urban <- as.numeric(as.character(pca_data$Urban))

pca_result <- prcomp(
  pca_data,
  center = TRUE,
  scale. = TRUE,
)

biplot(pca_result)

pc_county <- as.data.frame(pca_result$x[, 1:2])
colnames(pc_county) <- c("PC1", "PC2")

head(pc_county)

pca_result$rotation[,1]

```

The features were centered and scaled before running PCA because PCA is sensitive to variance of the input features. So features with large raw magnitudes would dominate if we dont normalize them.

The three highest absolute values for PC1 are pct_male, pct_female, and pct_whitealone. For PC1, pct_male and white alone has an opposite sign from the others. Negative sign indicates a features favorability towards the democratic candidate, and positive indicates favorability towards the republican. This means male and white alone are inversely corellated with everything else. This makes sense since the PC1 values of male and female pct are exactly the opposite sign from each other. We addressed the issue of Urban-Rural being qualitative by converting it to binary classification earlier.

11\.

```{r}
vars <- (pca_result$sdev)^2/sum((pca_result$sdev)^2)
cum_var <- cumsum(vars)
cum_var
```

We need at least 5 PCs to capture 90% of the variance for the analysis.

Plots:

```{r}
pve <- data.frame(
  PC = 1:length(vars),
  PVE = vars, 
  CumPVE = cum_var
)

ggplot(pve, aes(x=PC, y=PVE))+
  geom_col() + 
  labs(
    title = "Proportion of Variance Explained",
    x = "Prin. Comp.",
    y = "PVE"
  ) +
  theme_classic()
```

```{r}

ggplot(pve, aes(x=PC, y=CumPVE))+
  geom_col() + 
  labs(
    title = "Cumulative Proportion of Variance Explained",
    x = "Prin. Comp.",
    y = "Cum. PVE"
  ) +
  theme_classic()
```

12. Clustering

```{r}
cluster_data <- census_clean %>%
  select(-State, -County)

cluster_data$Urban <- as.numeric(as.character(cluster_data$Urban))

cluster_scaled <- scale(cluster_data)

hc_original <- hclust(dist(cluster_scaled), method = 'complete')

plot(
  hc_original,
  labels = FALSE,
  main = "Hierarchical Clustering Using Census Variables",
  xlab = " ",
  sub = " "
)

clusters_original <- cutree(hc_original, k = 10)

census_clusters_original <- census_clean %>%
  mutate(cluster_original = clusters_original)

santa_barbara_original <- census_clusters_original %>%
  filter(State == "California", County == "Santa Barbara County")

santa_barbara_original

```

```{r}
pc_scaled <- scale(pc_county)

hc_pc <- hclust(dist(pc_scaled), method = "complete")

plot(
  hc_pc,
  labels = FALSE,
  main = "Hierarchical Clustering Using PC1 and PC2",
  xlab = " ",
  sub = " "
)

clusters_pc <-cutree(hc_pc, k = 10)

census_clusters_pc <- census_clean %>%
  mutate(clusters_pc = clusters_pc)

head(census_clusters_pc)

santa_barbara_pc <- census_clusters_pc %>%
  filter(State == "California", County == "Santa Barbara County")

santa_barbara_pc
```

```{r}
sb_original_cluster <- santa_barbara_original$cluster_original

sb_original_counties <- census_clusters_original %>%
  filter(cluster_original == sb_original_cluster) %>%
  select(State, County, 'Median Household Income', Urban, pct_male, pct_female, pct_whitealone, pct_black, pct_hispanic)

head(sb_original_counties, 15)

sb_pc_cluster <- santa_barbara_pc$clusters_pc

sb_pc_counties <- census_clusters_pc %>%
  filter(clusters_pc == sb_pc_cluster) %>%
  select(State, County, 'Median Household Income', Urban, pct_male, pct_female, pct_whitealone, pct_black, pct_hispanic)

head(sb_pc_counties, 15)
```

The PCA based clustering placed Santa Barbara County in a more appropriate cluster. the original clustering grouped Santa Barbara with over 2000 counties while the PCA based clustering produced a smaller and more meaningful group. This most likely happened because PCA reduces noise and multicollinearity by focusing on the main sources of variation in the data.

## Classification

```{r}
# we move all state and county names into lower-case
# we further remove suffixes of "county"
tmpwinner <- county_winner %>% ungroup %>%
  mutate_at(vars(State, County), tolower) %>%
  mutate(County = gsub(" county", "", County)) 

# we move all state and county names into lower-case
# we further remove suffixes of "county"
tmpcensus <- census_clean %>% mutate_at(vars(State, County), tolower) %>%
  mutate(County = gsub(" county", "", County)) 

# we join the two datasets
election_cl <- tmpwinner %>%
  left_join(tmpcensus, by = c("State" = "State", "County" = "County")) %>% 
  na.omit

## save meta information
election_meta <- election_cl %>% select(c(State, FIPS, County, total_gop, total_dem))

## save predictors and class labels
election_cl = election_cl %>% select(-c(State, FIPS, County, total_gop, total_dem, pct_female))
set.seed(10) 
n <- nrow(election_cl)
idx.tr <- sample.int(n, 0.8*n) 
election.tr <- election_cl[idx.tr, ]
election.te <- election_cl[-idx.tr, ]
set.seed(20) 
nfold <- 10
folds <- sample(cut(1:nrow(election.tr), breaks=nfold, labels=FALSE))
calc_error_rate = function(predicted.value, true.value){
  return(mean(true.value!=predicted.value))
}
records = matrix(NA, nrow=3, ncol=2)
colnames(records) = c("train.error","test.error")
rownames(records) = c("tree","logistic","lasso")
```

13. decision tree

```{r}
names(election.tr)[names(election.tr) == "Median Household Income"] <-
  "Median_Household_Income"

names(election.te)[names(election.te) == "Median Household Income"] <-
  "Median_Household_Income"
election.tr$winner <- factor(election.tr$winner)
election.te$winner <- factor(election.te$winner)
str(election.tr$winner)
```

```{r}
library(tree)

tree_fit <- tree(
  winner ~ .,
  data = as.data.frame(election.tr)
)
cv_output <- cv.tree(
  tree_fit,
  FUN = prune.misclass,
  rand = folds
)

best_size <- cv_output$size[which.min(cv_output$dev)]


pruned_tree <- prune.misclass(
  tree_fit,
  best = best_size
)

train_pred <- predict(
  pruned_tree,
  election.tr,
  type = "class"
)

test_pred <- predict(
  pruned_tree,
  election.te,
  type = "class"
)

train_error <- mean(train_pred != election.tr$winner)
test_error  <- mean(test_pred  != election.te$winner)

records["tree", "train.error"] <- train_error
records["tree", "test.error"]  <- test_error


```

store

```{r}
records
```

interpretation

```{r}
pruned_tree
plot(pruned_tree)
text(pruned_tree, pretty = 0)
```

pct_white alone is at the top of the decision tree, meaning it is the top predictor. One of the most interesting things we found was that if the white percentage is greater than 50%, and the household income is greater than 83618, then the county is likely to vote democratic if it has less than 70% white people. This is basically describing California. Also, if the black percentage is greater than 54%, then the county will vote democratic without any conditions. Also, if it is less than that, having LESS than \$76423 household income means a county will vote GOP.

14. logistic regression

```{r}
log_fit <- glm(
  winner ~ .,
  data = election.tr,
  family = "binomial"
)
summary(log_fit)
```

In the training set, all of the variables are significant at 0.05. Urban, Income, and White are the most significant. White has the highest coefficient out of the most significant variables.

predictions (probabilities of GOP)

```{r}
train_prob <- predict(
  log_fit,
  newdata = election.tr,
  type = "response"
)
test_prob <- predict(
  log_fit,
  newdata = election.te,
  type = "response"
)
logtrain_pred <- ifelse(train_prob>0.5, "gop", "dem")
logtest_pred <- ifelse(test_prob>0.5, "gop", "dem")

logtrain_pred <- factor(
  logtrain_pred,
  levels = levels(election.tr$winner)
)

logtest_pred <- factor(
  logtest_pred,
  levels = levels(election.te$winner)
)

logtrain_error <- mean(
  logtrain_pred != election.tr$winner
)

logtest_error <- mean(
  logtest_pred != election.te$winner
)
records["logistic", "train.error"] <- logtrain_error
records["logistic", "test.error"]  <- logtest_error
```

```{r}
exp(coef(log_fit)*0.01)

```

A 1% increase in the white share of a county's population increases the odds of the county voting republican by 7.74%, keeping everything else constant. A 1% increase in the male share of a county's population increases the odds of the county voting republican by 8.31%, keeping everything else constant, but this is less significant. Probably because there is less variance in this group than in white population in different counties.

15\.

```{r}
library(glmnet)

x_train <- model.matrix(winner ~ ., data = election.tr)[, -1]
y_train <- election.tr$winner

x_test <- model.matrix(winner ~ ., data = election.te)[, -1]
y_test <- election.te$winner

y_train_num <- ifelse(y_train == "gop", 1, 0)
y_test_num <- ifelse(y_test == "gop", 1, 0)

lambda_grid <- seq(1, 50) * 1e-4

set.seed(20)
lasso_cv <- cv.glmnet(
  x = x_train,
  y = y_train_num,
  family = "binomial",
  alpha = 1,
  lambda = lambda_grid,
  nfolds = 10
)

best_lambda <- lasso_cv$lambda.min
best_lambda

plot(lasso_cv)

lasso_coef <- coef(lasso_cv, s = "lambda.min")
lasso_coef

nonzero_coef <- lasso_coef[lasso_coef[, 1] !=0, , drop = FALSE]
nonzero_coef

lasso_train_prob <- predict(lasso_cv, newx = x_train, s = "lambda.min", type = "response")
lasso_test_prob <- predict(lasso_cv, newx = x_test, s = "lambda.min", type = "response")

lasso_train_pred <- ifelse(lasso_train_prob > 0.5, "gop", "dem")
lasso_test_pred <- ifelse(lasso_test_prob > 0.5, "gop", "dem")

lasso_train_pred <- factor(lasso_train_pred, levels = levels(y_train))
lasso_test_pred <- factor(lasso_test_pred, levels = levels(y_test))

records["lasso", "train.error"] <- mean(lasso_train_pred != y_train)
records["lasso", "test.error"] <- mean(lasso_test_pred != y_test)

records

```

The optimal lambda from cross validation was 0.0001. All predictors remained in the LASSO model so non were removed. This suggests each variable contributed useful information for predicting the winning candidate. Logistic regression had the lowest test error (13.96%). Overall logistic regression performed the best and regularization did not improve accuracy for this dataset.

16. Compute ROC Curves

```{r}
library(pROC)

tree_prob <- predict(pruned_tree, newdata = election.te, type = "vector") [, "gop"]

log_prob <- predict(log_fit,
                    newdata = election.te,
                    type = "response")

lasso_prob <- predict(lasso_cv,
                      newx = x_test,
                      s = "lambda.min",
                      type = "response")

y_true <- ifelse(election.te$winner == "gop", 1, 0)

roc_tree <- roc(y_true, as.numeric(tree_prob))
roc_log <-roc(y_true, as.numeric(log_prob))
roc_lasso <-roc(y_true, as.numeric(lasso_prob))

plot(roc_tree,
     col = "blue",
     main = "ROC Curves",
     legacy.axes = TRUE)

plot(roc_log,
     col = "red",
     add = TRUE)

plot(roc_lasso,
     col = "green",
     add = TRUE)

legend("bottomright",
       legend = c(
         paste("Tree AUC =", round(auc(roc_tree),3)),
         paste("Logistic AUC =", round(auc(roc_log),3)),
         paste("LASSO AUC =", round(auc(roc_lasso),3))
       ),
       col = c("blue","red","green"),
       lwd = 2)

auc(roc_tree)
auc(roc_log)
auc(roc_lasso)
```

The ROC curves show that all three models perform well at predicting election outcomes. The Decision Tree had the highest AUC (0.862) then followed by Logistic AUC (0.851) and LASSO AUC (0.824). This indicated that the decion tree provided the best classification performance on the test data. However logistic regression and LASSO offer greater interpretability. With LASSO also helping reduce overfitting through regularization; though it has higher error than the other models. Overall the decision tree was the best for prediction while the regression based methods were better for understanding the effects of individual variables.

## Taking it Further

17. Explore additional methods

    We will try linear discriminant analysis and quadratic discriminant analysis.

    **LDA:**

    Fitting:

    ```{r}
    library(MASS)
    lda_fit <- lda(
      winner ~ .,
      data = election.tr
    )
    ```

Prediction:

```{r}
lda <- c(0, 0)
records <- rbind(records, lda)
lda_train_pred <- predict(lda_fit, election.tr)$class
lda_test_pred <- predict(lda_fit, election.te)$class
records["lda", "train.error"] <- mean(lda_train_pred != election.tr$winner)
records["lda", "test.error"] <- mean(lda_test_pred != election.te$winner)

```

LDA and Lasso are weaker than tree and logistic regression in error. This weakness suggests that the data may not fully fulfill the LDA assumptions of normal predictors and equal covariance between predictors.

**QDA:**

Fitting:

```{r}
qda_fit <- qda(
  winner ~ .,
  data = election.tr
)

```

Pred:

```{r}
qda <- c(0, 0)
records <- rbind(records, qda)
lda_train_pred <- predict(lda_fit, election.tr)$class
lda_test_pred <- predict(lda_fit, election.te)$class
records["lda", "train.error"] <- mean(lda_train_pred != election.tr$winner)
records["lda", "test.error"] <- mean(lda_test_pred != election.te$winner)
```
